---
title: "Chapter 2: Linear Regression"
author: "Gyeong min Kim"
date: November 19, 2024
institute: Department of Statistics \newline Sungshin Women’s University
fonttheme: "serif"
fontsize: 8pt
output:
  beamer_presentation:
    latex_engine: xelatex 
    theme: "metropolis"
header-includes:
  - \input{header_includes.tex}
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(fig.height = 4, fig.width = 6)
set.seed(1)
```




# Distribution of the RSS Values

## Hat matrix $H$


- We explore the properties of the matrix
$$
H \stackrel{\triangle}{=} X (X^\top X)^{-1} X^\top \in \R^{n \times n}.
$$

-  The following are easy to derive but useful in the later part of this book:
\begin{align*}
\blue{H^2} &=  X (X^\top X)^{-1} X^\top X (X^\top X)^{-1} X^\top = X (X^\top X)^{-1} X^\top = \blue{H} \\
\blue{(I - H)^2} &= I -2H + H^2 = I - 2H + H = \blue{I - H} \\
\blue{HX} &= X (X^\top X)^{-1} X^\top X = \blue{X}.
\end{align*}


- Moreover, if we set $\hat y = X \hat \beta$, we have 
$$
\hat y = X \hat \beta = X (X^\top X)^{-1} X^\top y = Hy.
$$

- And we observe
\begin{align*}
y - \hat y &= (I - H)y = (I - H)(X \beta + \e) \\
           &= X\beta + \e - HX\beta -H\e = X\beta + \e - X\beta -H\e\\
           &= (I - H) \e.
\end{align*}



## The RSS with respect to Hat matrix $H$

- Observe the equation
\begin{align*}
RSS = \|y - \hat y  \|^2 = \e^\top (I - H)^\top (I - H) \e = \e^\top (I - H)^2 \e = \e^\top (I - H) \e. 
\end{align*}


- To analysis $RSS$, we explore the properties of Hat matrix $H$.


\begin{proposition} \label{I-H diagonalization}
If $\rank(X) = p + 1$, we obtain the diagonalization
$$
P^\top (I - H) P = \diag( \underbrace{1, ..., 1}_{N-p-1}, \underbrace{0, ..., 0}_{p + 1} ),
$$
where $P$ is orthonormal matrix whose columns consist of eigenvectors of matrix $I - H.$ 
\end{proposition}




## Proof of Proposition \ref{I-H diagonalization}

- If $\rank(X) = p + 1$, we have
\begin{align*}
\rank(H) &= \rank\left(X(X^\top X)^{-1} \cdot X^\top  \right) \\
         &\leq \min \left\{ \rank(X(X^\top X)^{-1}), \rank(X) \right\} \\
         &\leq \rank(X) = p + 1
\end{align*}


- If $\rank(X) = p + 1$, we have
\begin{align*}
\rank(H) &\geq \min \left\{ \rank(H), \rank(X) \right\} \\
         &\geq \rank(HX) = \rank(X) = p + 1
\end{align*}


- We conclude that \navy{$\mathbf{\rankbf(X) = p + 1 \quad \Rightarrow \quad \rankbf(H) = p + 1}$}.


## Proof of Proposition \ref{I-H diagonalization} (cntd.)

- Recall the relationship $HX = X$:
$$
HX = H\BM | & | &   & | \\
          X_1 & X_2 & \dots & X_{p+1} \\
          | & | &   & |\EM
   = \BM | & | &   & | \\
          X_1 & X_2 & \dots & X_{p+1} \\
          | & | &   & |\EM .
$$

- We have
$$
HX_i = X_i \for i = 1, ..., p+1.
$$

- $\rank(X) = p + 1 \quad \Rightarrow \quad \dim(\Eigen(H)) = p + 1$ where $\Eigen(H)$ denotes the eigenspace with eigenvectors corresponding non-zero eigenvalue.


- Therefore, each column of $X$ spans the eigenspace of $H$, which means \navy{$X_i$s are the eigenvectors of the Hat matrix $H$}.




## Proof of Proposition \ref{I-H diagonalization} (cntd.)

- Now, we analyze the relationship between $\Eigen(H)$ and $\mathcal{N}(I-H)$ where $\mathcal{N}$ denotes the nullspace of a matrix.


- For arbitray $x \in \R^n$,
$$
Hx = x \quad \Rightarrow \quad (I-H)x = \mathbf{0},
$$
which means that the eigenvectors of $H$ belong to the nullspace of $I-H$. 


- For arbitray $x \in \R^n$,
$$
(I-H)x = \mathbf{0} \quad \Rightarrow \quad Hx = x ,
$$
which means that the vectors of $\mathcal N (I-H)$ belong to the $\Eigen(H)$. 


- Therefore, we have $\mathcal N (I-H) = \Eigen(H)$, and 
$$
\mathcal \dim \mathcal N (I-H) = \dim \Eigen(H) = p + 1.
$$


- Then we observe $\rank (I-H) = n - p - 1$


## Proof of Proposition \ref{I-H diagonalization} (cntd.)


- $I-H$ is diagonalizable matrix, since it is the symmetric and square matrix.

- Since $\rank(I-H) = n - p - 1$, we have
$$
P^\top(I - H)P = \diag(\underbrace{\lambda_1, ..., \lambda_{n-p-1}}_{n-p-1}, \underbrace{0, ..., 0}_{p + 1}),
$$
where columns of $P$ are orthonormal and consist of eigenvectors of $I-H$.









## Proof of Proposition \ref{I-H diagonalization} (cntd.)

\begin{lemma} \label{eigenvalue}
Let a real matrix $D \in \R^{n \times n}$ such that $D^2 = D$. Then, the eigenvalues of $D$ consist of only $0$ and $1$.
\end{lemma}
\begin{proof}
Let $D \in \R^{n \times n}$ such that $D^2 = D$, and
$$
\exists v \in \R^n, \,\,\, Dv = \lambda v.
$$
Then, we observe
\begin{align*}
Dv = \lambda v \quad \Rightarrow \quad D^2v = \lambda D v \quad \Rightarrow \quad Dv = \lambda^2 v,
\end{align*}
since $D^2 = D$. Thus, we have 
$$
\lambda = \lambda^2 \quad \Rightarrow \quad \lambda = 0 \,\, \text{or} \,\, 1.
$$
\end{proof}




## Proof of Proposition \ref{I-H diagonalization} (cntd.)

- In order to proof Proposition \ref{I-H diagonalization}, we apply the following:

  1. $\rank(X) = p+1 \quad \Rightarrow \quad \rank(H) = p + 1$
  
  2. $\mathcal \dim \mathcal N (I-H) = \dim \Eigen(H) = p + 1 \quad \Rightarrow \quad \rank(I-H) = n -p - 1$
  
  3. $P^\top(I - H)P = \diag(\underbrace{\lambda_1, ..., \lambda_{n-p-1}}_{n-p-1}, \underbrace{0, ..., 0}_{p + 1})$
  
  4. $(I - H)^2 = (I - H) \quad \Rightarrow \quad \lambda = 0\,\, \text{or} \,\, 1$.
  

- Therefore, if $\rank(X) = p+1$, we obtain
$$
P^\top (I - H) P = \diag( \underbrace{1, ..., 1}_{n-p-1}, \underbrace{0, ..., 0}_{p + 1} ).
$$


## Distribution of RSS values

- Since the columns of $P$ are orthonormal,
$$
\exists u \in \R^n, \,\,\, \e = Pu, \quad \text{and then} \quad u = P^\top \e.
$$

- We have
\begin{align*}
RSS &= \e^\top (I - H) \e = u^\top P^\top (I-P) P u \\
    &= u^\top \diag( \underbrace{1, ..., 1}_{n-p-1}, \underbrace{0, ..., 0}_{p + 1} ) u \\
    &= \sum_{i = 1}^{n - p -1} u_i^2.
\end{align*}



- We observe 
\begin{align*}
\E(u) &= \E(P\e) = \textbf{0} \\
\COV(u) &= \E(uu^\top) = P^\top E(\e\e^\top) P = \sigma^2 P^\top P = \sigma^2 I_n. 
\end{align*}


- Since $u \sim N_n(0, \sigma^2 I_n)$ and then $\frac{u}{\sigma} \sim N_n(0, I_n)$,
$$
\frac{RSS}{\sigma^2} = \sum_{i = 1}^{n - p -1} \left( \frac{u_i}{\sigma} \right)^2 \sim \mathcal{X}^2_{n-p-1}.
$$


## Plot of Chi-squared distribution
```{r}
i = 1 ; curve(dchisq(x, i), 0, 20, col = i)
for(i in 2:10) curve(dchisq(x, i), 0, 20, col = i, add = TRUE, ann = FALSE)
legend("topright", legend = 1:10, lty = 1, col = 1:10)
```


# Hypothesis Testing for $\hat \beta_j \neq 0$

## Distribution of $\hat \beta$

- In this section, we consider whether each of the $\beta_j,\,\,j = 0, 1,..., p,$ is zero or not based on the data.


- Due to fluctuations in the $N$ random variables $\e_1, ..., \e_n$, the data occurred by chance.


- Since the estimator $\hat \beta = (X^\top X)^{-1} X^\top y$ have randomness, we use the distribution of the estimator:


$$
\hat \beta \sim N_{p+1}\left( \beta,\,\, \sigma^2 (X^\top X)^{-1} \right).
$$

## Known or Unknwon $\sigma^2$

- If we know $\sigma^2$, use the $z$-statistics
$$
z = \frac{\hat \beta_j -\beta_j}{\sqrt{ \VAR(\hat\beta_j)}} = \frac{\hat \beta_j -\beta_j}{\sigma \sqrt{(X^\top X)^{-1}_{jj}}} \sim N(0, 1).
$$

- If we do not know $\sigma^2$, use the $t$-statistics
$$
t = \frac{\hat \beta_j -\beta_j}{\sqrt{\widehat{\VAR}(\hat\beta_j)}} = \frac{\hat \beta_j -\beta_j}{\hat \sigma\sqrt{(X^\top X)^{-1}_{jj}}} \sim t_{n - p - 1}. 
$$


## Unknwon $\sigma^2$

- When $\sigma^2$ is unknown, we need to estimate $\sigma^2$. Since $\frac{RSS}{\sigma^2} \sim \mathcal{X}^2_{n-p-1}$,
$$
\E\left( \frac{RSS}{n - p - 1} \right) = \sigma^2,
$$
where $RSS/(n-p-1)$ is unbiased estimator of $\sigma^2$


$$ 
\left( \therefore \,\, \hat \sigma^2 = \frac{RSS}{n - p - 1} \right)
$$

## Unknwon $\sigma^2$

- Let $U \sim N(0, 1)$ and $V \sim \mathcal X_{df}$, then
$$
\frac{U}{\sqrt{V/{df}}} \sim t_{df}.
$$


- $t$-statistics
\begin{align*}
t &= \frac{\hat \beta_j -\beta_j}{\hat \sigma \sqrt{(X^\top X)^{-1}_{jj}}} \\
  &= \frac{\hat \beta_j -\beta_j}{\sigma \sqrt{(X^\top X)^{-1}_{jj}}}  \sqrt \frac{\sigma^2}{\hat \sigma^2}
   = \frac{\hat \beta_j -\beta_j}{\sigma \sqrt{(X^\top X)^{-1}_{jj}}} / \sqrt \frac{\hat \sigma^2}{\sigma^2}\\
  &= \frac{\hat \beta_j -\beta_j}{\sigma \sqrt{(X^\top X)^{-1}_{jj}}} / \sqrt{ \frac{RSS}{\sigma^2} / (n - p - 1) } \\
  &= \frac{U}{\sqrt{V/{df}}} \sim t_{df} = t_{n - p - 1}
\end{align*}


## [Example 1] Under $H_0: \beta_j = 0$
```{r}
n = 100 ; p = 1 ; rep = 1000
T = NULL
t = rep(0, rep)
for(i in 1:rep){
  x = rnorm(n) ; y = rnorm(n) 
  fit = lm(y ~ x)
  RSS = crossprod(y - fit$fitted.values) 
  sigma_hat = sqrt( RSS / (n - p - 1) )
  statistics = fit$coefficients[2] / ( sigma_hat / sqrt(crossprod(x - mean(x))) )
  t[i] = statistics
}
```


## Plot of Example 1
```{r echo = FALSE}
hist(t, breaks = sqrt(rep), probability=TRUE,
     xlab="t", ylab="Probability Density Function",
     main="Histogram of the value of t and its theoretical distribution in red")
curve(dt(x, n - p - 1), -3, 3, type="l", col = "red", add = TRUE)
```


## [Example 2] Not under $H_0: \beta_j = 0$
```{r}
n = 100 ; p = 1 ; rep = 1000
T = NULL
t = rep(0, rep)
for(i in 1:rep){
  x = rnorm(n) ; y = 0.1*x + rnorm(n) 
  fit = lm(y ~ x)
  RSS = crossprod(y - fit$fitted.values) 
  sigma_hat = sqrt( RSS / (n - p - 1) )
  statistics = fit$coefficients[2] / ( sigma_hat / sqrt(crossprod(x - mean(x))) )
  t[i] = statistics
}
```


## Plot of Example 2
```{r echo = FALSE}
hist(t, breaks = sqrt(rep), probability=TRUE,
     xlab="t", ylab="Probability Density Function",
     main="Histogram of the value of t and its theoretical distribution in red")
curve(dt(x, n - p - 1), -3, 3, type="l", col = "red", add = TRUE)
```


# Coefficient of Determination and the Detection of Collinearity

## Three sum of squares

- Let $W \in \R^{n \times n}$ be a matrix such that all the elements are $1/n$. Then we have $Wy = (\bar y, ..., \bar y )\in \R^n$

- Total sum of squares (TSS):
$$
TSS \stackrel{\triangle}{=} \| y - \bar y \cdot \mathbf{1}  \|_2^2 
                         =  \| y - W y \|_2^2 
                         =  \| (I - W )y \|_2^2 
$$

- Residual sum of squares (RSS):
$$
RSS \stackrel{\triangle}{=} \| y - \hat y \|_2^2 
                         =  \| y - H y \|_2^2 
                         =  \| (I - H )y \|_2^2 
$$

- Explained sum of squares (ESS):
$$
ESS \stackrel{\triangle}{=} \| \hat y - \bar y \cdot \mathbf{1}  \|_2^2 
                         =  \| H y - W y \|_2^2 
                         =  \| (H - W )y \|_2^2 
$$


## Total sum of squares Decomposition

\begin{align*}
TSS = 
\end{align*}



























