---
title: "Chapter 2: Linear Regression"
author: "Gyeong min Kim"
date: November 19, 2024
institute: Department of Statistics \newline Sungshin Women’s University
fonttheme: "serif"
fontsize: 8pt
output:
  beamer_presentation:
    latex_engine: xelatex 
    theme: "metropolis"
header-includes:
  - \input{header_includes.tex}
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(fig.height = 4, fig.width = 6)
set.seed(1)
```


# Least Squares Method for Simple Linear Regression
## Data Generation
```{r}
beta = c(-0.5, 1)
n = 100 ; x = rnorm(n, mean = 1) ; y = beta[1] + beta[2] * x + rnorm(n)
plot(x, y)
```


## Least Squares algorithm for Simple Linear Regression
```{r}
ls = function(x, y){
  beta_hat1 = crossprod(x - mean(x), y - mean(y)) / crossprod(x - mean(x))
  beta_hat0 = mean(y) - beta_hat1 * mean(x)
  
  return(list("intercept" = as.numeric(beta_hat0),
              "slope" = as.numeric(beta_hat1)))
}

beta ; ls(x, y)
```

## Plot of Simple Linear regression (Original vs. Centering)
```{r}
c(ls(x, y)$intercept, ls(x - mean(x), y - mean(y))$intercept)
c(ls(x, y)$slope, ls(x - mean(x), y - mean(y))$slope)
```
```{r echo=FALSE, fig.height=5}
plot(x, y) ; abline(h = 0) ; abline(v = 0)
abline(ls(x, y)$intercept, ls(x, y)$slope, col = "red")
abline(ls(x - mean(x), y - mean(y))$intercept, 
       ls(x - mean(x), y - mean(y))$slope, col = "blue")
legend("bottomright", c("BEFORE", "AFTER"), lty = 1, col=c("red", "blue"))
```

## Plot of Simple Linear regression (Original vs. Centering)*
```{r}
rbind("Original" = ls(x, y), 
      "Centerd X" = ls(x - mean(x), y),
      "Centerd Y" = ls(x, y - mean(y)),
      "Centerd X and Y" = ls(x - mean(x), y - mean(y)))
```


```{r echo=FALSE, fig.height=5}
plot(x, y) ; abline(h = 0) ; abline(v = 0)
abline(ls(x, y)$intercept, ls(x, y)$slope, col = "red")
abline(ls(x - mean(x), y)$intercept, ls(x - mean(x), y)$slope, col = "orange")
abline(ls(x, y - mean(y))$intercept, ls(x, y - mean(y))$slope, col = "yellow2")
abline(ls(x - mean(x), y - mean(y))$intercept, ls(x - mean(x), y - mean(y))$slope, col = "green3")
legend("bottomright", c("Original", "Centerd X", "Centerd Y", "Centerd X and Y"),
       lty = 1, col=c("red", "orange", "yellow2", "green3"))
```

# Least Squares Method for Multiple Linear Regression

## Multiple Linear Regression scheme

- Consider the multiple linear regression:
$$
y = X \beta + \varepsilon.
$$
  - $\varepsilon = \left( \begin{matrix} \varepsilon_1 \\ \varepsilon_2 \\ \vdots \\ \varepsilon_n \end{matrix} \right) \in \mathbf R^n \where \varepsilon_1, \varepsilon_2, ..., \varepsilon_n \stackrel{iid}{\sim} N(0, \sigma^2)$.
  \vt
  - $y = \left( \begin{matrix} y_1 \\ y_2 \\ \vdots \\ y_n \end{matrix} \right) \in \mathbf R^n, \,\,\, X = \left[ \begin{matrix} 1&x_{11}&\dots&x_{1p} \\ 1&x_{21}&\dots&x_{2p} \\ \vdots & \vdots & \ddots&\vdots \\ 1&x_{n1}&\dots&x_{np} \end{matrix} \right]\in \mathbf R^{n \times (p+1)}, \,\,\, \text{and}\,\,\, \beta = \left( \begin{matrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_p \end{matrix} \right)\in \mathbf R^{p+1}$.
  

## Solution of Multiple Linear Regression

- When a matrix $X^\top X \in \mathbf R^{(p + 1) \times (p + 1)}$ is invertible, we have
$$
\hat \beta = \left( X^\top X \right)^{-1} X^\top y.
$$



## Data Generation and Least Squares Estimator
```{r}
beta = c(1, 2, 3)
n = 100 ; p = 3
X = cbind("intercept" = 1, "x1" = rnorm(n), "x2" = rnorm(n))
y = X %*% beta + rnorm(n)
solve(t(X) %*% X) %*% t(X) %*% y # Original


C = cbind(1, X[,2] - mean(X[,2]), X[,3] - mean(X[,3]))
solve(t(C) %*% C) %*% t(C) %*% (y - mean(y)) # Centered X and Y
```



## Rank Condition of Design matrix $X$


- We may notice that the matrix $X^\top X$ is not invertible under each of the following conditions:
  
  1. $N < p + 1$
  
  2. Two columns in $X$ coincide.
  


# Distribution of $\hat \beta$

## Moments of the estimator $\hat \beta$
$$
\hat \beta = \left( X^\top X \right)^{-1} X^\top y
$$



- The estimate $\hat \beta$ of $\beta$ depends on the value of $\e$ because $N$ pairs of data $(x_1, y_1),..., (x_n, y_n )$ randomly occur. 


\begin{align*}
\E \left( \hat \beta\right) &= \left( X^\top X \right)^{-1} X^\top \E \left(y\right) 
                             = \left( X^\top X \right)^{-1} X^\top X \beta = \beta \\
\var \left( \hat \beta\right)
&= \left( X^\top X \right)^{-1} X^\top \var(\e) X \left( X^\top X \right)^{-1}                              = \sigma^2 \left( X^\top X \right)^{-1}
\end{align*}

$$
\left( \therefore \,\, \hat \beta \sim N(\beta, \sigma^2 \left( X^\top X \right)^{-1} )\right)
$$



# Distribution of the RSS Values

## Hat matrix $H$


- We explore the properties of the matrix
$$
H \stackrel{\triangle}{=} X (X^\top X)^{-1} X^\top \in \R^{n \times n}.
$$

-  The following are easy to derive but useful in the later part of this book:
\begin{align*}
\blue{H^2} &=  X (X^\top X)^{-1} X^\top X (X^\top X)^{-1} X^\top = X (X^\top X)^{-1} X^\top = \blue{H} \\
\blue{(I - H)^2} &= I -2H + H^2 = I - 2H + H = \blue{I - H} \\
\blue{HX} &= X (X^\top X)^{-1} X^\top X = \blue{X}.
\end{align*}


- Moreover, if we set $\hat y = X \hat \beta$, we have 
$$
\hat y = X \hat \beta = X (X^\top X)^{-1} X^\top y = Hy.
$$

- And we observe
\begin{align*}
y - \hat y &= (I - H)y = (I - H)(X \beta + \e) \\
           &= X\beta + \e - HX\beta -H\e = X\beta + \e - X\beta -H\e\\
           &= (I - H) \e.
\end{align*}



## The RSS with respect to Hat matrix $H$

- Observe the equation
\begin{align*}
RSS = \|y - \hat y  \|^2 = \e^\top (I - H)^\top (I - H) \e = \e^\top (I - H)^2 \e = \e^\top (I - H) \e. 
\end{align*}


- To analysis $RSS$, we explore the properties of Hat matrix $H$.


\begin{proposition} \label{I-H diagonalization}
If $\rank(X) = p + 1$, we obtain the diagonalization
$$
P (I - H) P^\top = \diag( \underbrace{1, ..., 1}_{N-p-1}, \underbrace{0, ..., 0}_{p + 1} ),
$$
where $P$ is orthonormal matrix whose columns consist of eigenvectors of matrix $I - H.$ 
\end{proposition}




## Proof of Proposition \ref{I-H diagonalization}

- If $\rank(X) = p + 1$, we have
\begin{align*}
\rank(H) &= \rank\left(X(X^\top X)^{-1} \cdot X^\top  \right) \\
         &\leq \min \left\{ \rank(X(X^\top X)^{-1}), \rank(X) \right\} \\
         &\leq \rank(X) = p + 1
\end{align*}


- If $\rank(X) = p + 1$, we have
\begin{align*}
\rank(H) &\geq \min \left\{ \rank(H), \rank(X) \right\} \\
         &\geq \rank(HX) = \rank(X) = p + 1
\end{align*}


- We conclude if $\rank(X) = p + 1$, \blue{$\rank(H) = p + 1$}.




## Proof of Proposition \ref{I-H diagonalization} (cntd.)

- Recall the relationship $HX = X$:
$$
HX = H\BM | & | &   & | \\
          X_1 & X_2 & \dots & X_{p+1} \\
          | & | &   & |\EM
   = \BM | & | &   & | \\
          X_1 & X_2 & \dots & X_{p+1} \\
          | & | &   & |\EM 
$$

- For $i$-th column, $i=1, ..., p+1$, we have
$$
HX_i = X_i.
$$


























